Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science

### 2025-02-19 analysis/paper/paper.qmd
final polishing before submission

---
title: "Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science"
format: html
bibliography: references.bib
fig-dpi: 600
execute: 
    echo: false
    warning: false
---

```{=html}
<!--

authoritative, agenda-setting or hypothesis-setting papers, rather than descriptive compilations of the literature.

have the vision and authority to guide the next generation of archaeological science research

cross-ref Vaiglova 2025 this issue

-->
```

@@

# Number of authors ------------------
boxlplot_n_authors <- 
items_df %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             log(authors_n))) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 1.4,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 1.4,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.8,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.0,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.0,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 5)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal() + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of authors (ln)") +
  xlab("Collaborator group size")

@@

boxlplot_rel_title_length <- 
items_df_title %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             relative_title_length)) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.8,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.0,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.0,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = -0.2,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = -0.2,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0,
                  nudge_x = 0.05) +

  scale_y_continuous(limits = c(-4.5, 3),
                     breaks =  seq(-5, 5, 1),
                     labels = seq(-5, 5, 1)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal() +
  theme(panel.grid  = element_blank()) +
  ylab("Ratio of title length to article length (ln)") +
  xlab("Relative title length")

# Number of pages ------------------
boxlplot_n_pages <- 
items_df %>% 
  ggplot(aes(1,
             log(pages_n))) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 1.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 1.8,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 2.5,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 2.5,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 2.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 2.8,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_reverse(limits = c(5, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal() + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of pages (ln)") +
  xlab("Article length")

@@

# plot
boxlplot_price_index <- 
items_df %>% 
  ggplot(aes(1,
             prices_index)) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.38,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.38,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.29,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.29,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.19,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.19,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal() +
  theme(panel.grid  = element_blank()) +
  ylab("Prop. refs in last 5 years") +
  xlab("Recency of references")

@@

# plot
boxlplot_shannon_index <- 
shannon_per_item %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             shannon)) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 2.6,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 2.6,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 3.2,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 3.2,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 3.3,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 3.3,
                                label = "h"),
                  aes(x, y, 
                      label = label,
                      segment.colour = NA),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0,
                  nudge_x = 0.05) +
  scale_y_reverse(limits = c(6, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal()  +
  theme(panel.grid  = element_blank()) +
  ylab("Shannon Index") +
  xlab("Diversity of references")
```

```{r}
#| label: fig-compare-other-fields
#| fig-cap: Distributions of article characteristics hypothesised to reflect the level of consensus. The boxplot shows the distribution of values of archaeology articles. The thick black line in the middle of the boxplot is the median value, the box represents the inter-quartile range (the range between the 25th and 75th percentiles, where 50% of the data are located), and individual points represent outliers. The thin grey horizontal lines in each boxplot indicate the median values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities. ln denotes the natural logarithm, or logarithm to the base e. 
#| fig-height: 8

@@

```{r}
#| label: fig-change-over-time
#| fig-cap: Distribution of article characteristics for archaeology articles over time. Data points represent individual articles. The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series. 
#| fig-height: 6

@@

ggplot() + 
  geom_point(data = over_time_long_colour_gams_summary_df, 
       aes(year,  
           value, 
           colour = becoming_more_scientific),
       alpha = 0.5) + 
  geom_smooth(data = over_time_long_colour_gams_summary_df,
              aes(year, value),
              method="gam", 
              formula = y ~ s(x, bs = "cs"),
              se = FALSE,  
              size = 3, 
              colour = "#7570b3") +
  facet_wrap( ~ variable,
              scales = "free_y") + 
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("#d95f02",
                                "#1b9e77" )) +
  guides(colour = "none") +
  ylab("") +
  geom_text(data = over_time_long_colour_gams_summary_df %>% 
              group_by(variable) %>% 
              summarise(max_value = max(value),
                        adj.r.squared = unique(adj.r.squared)),
           aes(
           x = 1980, 
           y = max_value, 
           label = paste("Pseudo R² = ", 
                         signif(adj.r.squared, 
                                digits = 3))),
           hjust = 0, vjust = 1.5)

@@

journal_title_size <- 4

@@

# Plot the PCA results
plot_pca_means <- 
ggplot() + 
  labs(x = paste0("PC1 (", variance_explained[1], "%)"),
       y = paste0("PC2 (", variance_explained[2], "%)")) +
  geom_point(data = pca_scores_means,
               aes(.fittedPC1,
                   .fittedPC2),
             size = 2) +
  geom_text_repel(data = pca_scores_means,
            aes(.fittedPC1,
                .fittedPC2,
                label = journal),
            lineheight = 0.8,
            size = 2,
            bg.color = "white",  # Color of the halo
            bg.r = 0.2,) +
     geom_segment(data = pca_loadings_means, 
               aes(x = 0, 
                   y = 0, 
                   xend = PC1, 
                   yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "grey70") +
  geom_text_repel(data = pca_loadings_means,
            aes(PC1,
                PC2,
                label = column),
            size = 2,
            lineheight = 0.7,
            force = 0,
            segment.color = NA,
            color = "grey40",
           bg.color = "white",  # Color of the halo
          bg.r = 0.2) +
  theme_minimal()

@@

plot_borda_scores <- 
  borda_scores %>% 
  mutate(Journal = str_wrap(Journal, 20)) %>% 
ggplot() +
  aes(reorder(Journal, Score),
      Score) +
  geom_col() +
  coord_flip() +
  ylab("Borda Count scores") + 
  xlab("") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = journal_title_size))

@@

plot_journals_authors <- 
journal_metrics_for_plotting %>% 
  mutate(journal = str_wrap(journal, 20)) %>% 
  ggplot(aes(y = reorder(journal, 
                         log_authors,
                     FUN = mean),
             x = log_authors,
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  ylab("") + 
  xlab("Number of authors (ln)")

plot_journals_article_length <- 
journal_metrics_for_plotting %>% 
  mutate(journal = str_wrap(journal, 20)) %>% 
  ggplot(aes(y = reorder(journal,
                     -log_pages,
                     FUN = mean), 
             x = log_pages,
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  xlab("Number of pages (ln)") +
  ylab("")

plot_journals_title_length <- 
journal_metrics_for_plotting %>% 
  mutate(journal = str_wrap(journal, 20)) %>% 
  ggplot(aes(y = reorder(journal,
                         relative_title_length,
                         FUN = mean), 
             x = relative_title_length,
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  ylab("") +
  xlab("Relative title length (ln)")

plot_journals_ref_recency <- 
journal_metrics_for_plotting %>% 
  mutate(journal = str_wrap(journal, 20)) %>% 
  group_by(journal) %>% 
  ggplot(aes(y = reorder(journal,
                     prices_index,
                     FUN = mean), 
             x = prices_index,            
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  ylab("") +
  xlab("Recency of references")

plot_journals_ref_diversity <- 
journal_metrics_for_plotting %>%  
  mutate(journal = str_wrap(journal, 20)) %>% 
  group_by(journal) %>% 
  ggplot(aes(y = reorder(journal,
                     -shannon,
                     FUN = mean), 
             x = shannon,
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal() +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  ylab("") +
  xlab("Diversity of references")
```

```{r}
#| label: fig-variation-by-journal
#| fig-cap: "Panels A-E: Variation in bibliometric indicators of hardness for 20 archaeological journals. The journals are ordered for each indicator so that within each plot, the harder journals are at the top of the plot and the softer journals are at the base. Panel F shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm. Panel G is a biplot of the first and second principal components of a PCA computed on the journal means of the five bibliometric variables."
#| fig-height: 14

@@

plot_grid(
  plot_grid(plot_journals_authors, 
           plot_journals_article_length,
           plot_journals_title_length,
           plot_journals_ref_recency, 
           plot_journals_ref_diversity,
           plot_borda_scores,
           nrow = 2,
           labels = LETTERS[1:6]),
  plot_pca_means,
  nrow = 2,
  labels = c("", LETTERS[7]),
  rel_heights = c(1, 0.5))

@@

base_size <- 10

@@

# relationship between code issues and language?
aer_plot_issues_lang <- 
jas_rr_data %>% 
  separate_longer_delim(`Code issue`, ",") %>% 
  separate_longer_delim(Language, ",") %>% 
  mutate(Language = str_squish(Language)) %>% 
  mutate(`Code issue` = str_squish(`Code issue`)) %>% 
  group_by(`Code issue`, 
           Language) %>% 
  tally() %>% 
  drop_na() %>% 
  ggplot() +
  aes(`Code issue`, 
      Language,
      size = n) +
  geom_point() +
  xlab("Problem with the code") +
  ylab("Software") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  theme_minimal(base_size = base_size) +
  scale_size_area(breaks = c(1, 2, 3, 5)) +
  theme(legend.position = c(.9,.7),
  legend.box.background=element_rect(fill="white", 
                                     color="black")) 
```

>>>>

---
title: "Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Computational reproducibility is increasingly essential for advancing archaeological science, yet its implementation remains inconsistent across the discipline. This study presents a bibliometric assessment of archaeology’s status relative to other fields alongside a systematic review of computational reproducibility in published archaeological research. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

base_size <- 6

# Number of authors ------------------
boxlplot_n_authors <- 
items_df %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             log(authors_n))) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 1.4,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 1.4,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.8,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.0,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.0,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 5)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of authors (ln)") +
  xlab("Collaborator group size")

@@

boxlplot_rel_title_length <- 
items_df_title %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             relative_title_length)) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.8,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.0,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.0,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = -0.2,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = -0.2,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0,
                  nudge_x = 0.05) +
  scale_y_continuous(limits = c(-4.5, 3),
                     breaks =  seq(-5, 5, 1),
                     labels = seq(-5, 5, 1)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) +
  theme(panel.grid  = element_blank()) +
  ylab("Ratio of title length to article length (ln)") +
  xlab("Relative title length")

# Number of pages ------------------
boxlplot_n_pages <- 
items_df %>% 
  ggplot(aes(1,
             log(pages_n))) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 1.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 1.8,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 2.5,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 2.5,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 2.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 2.8,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_reverse(limits = c(5, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of pages (ln)") +
  xlab("Article length")

@@

# plot
boxlplot_price_index <- 
items_df %>% 
  ggplot(aes(1,
             prices_index)) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.38,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.38,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.29,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.29,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.19,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.19,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) +
  theme(panel.grid  = element_blank()) +
  ylab("Prop. refs in last 5 years") +
  xlab("Recency of references")

@@

# plot
boxlplot_shannon_index <- 
shannon_per_item %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             shannon)) +
  geom_boxplot(
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 2.6,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 2.6,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 3.2,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 3.2,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 3.3,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 3.3,
                                label = "h"),
                  aes(x, y, 
                      label = label,
                      segment.colour = NA),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0,
                  nudge_x = 0.05) +
  scale_y_reverse(limits = c(6, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size)  +
  theme(panel.grid  = element_blank()) +
  ylab("Shannon Index") +
  xlab("Diversity of references")
```

```{r}
#| label: fig-compare-other-fields
#| fig-cap: Distributions of article characteristics hypothesised to reflect the level of consensus. The boxplot shows the distribution of values of archaeology articles. The thick black line in the middle of the boxplot is the median value, the box represents the inter-quartile range (the range between the 25th and 75th percentiles, where 50% of the data are located), and individual points represent outliers. The thin grey horizontal lines in each boxplot indicate the median values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities. ln denotes the natural logarithm, or logarithm to the base e. 
#| fig-height: 4

@@

```{r}
#| label: fig-change-over-time
#| fig-cap: Distribution of article characteristics for archaeology articles over time. Data points represent individual articles. The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series. 
#| fig-height: 4

@@

ggplot() + 
  geom_point(data = over_time_long_colour_gams_summary_df, 
       aes(year,  
           value, 
           colour = becoming_more_scientific),
       alpha = 0.5) + 
  geom_smooth(data = over_time_long_colour_gams_summary_df,
              aes(year, value),
              method="gam", 
              formula = y ~ s(x, bs = "cs"),
              se = FALSE,  
              size = 2, 
              colour = "#7570b3") +
  facet_wrap( ~ variable,
              scales = "free_y") + 
  theme_bw(base_size = base_size) +
  scale_color_manual(values = c("#d95f02",
                                "#1b9e77" )) +
  guides(colour = "none") +
  ylab("") +
  geom_text(data = over_time_long_colour_gams_summary_df %>% 
              group_by(variable) %>% 
              summarise(max_value = max(value),
                        adj.r.squared = unique(adj.r.squared)),
           aes(
           x = 1980, 
           y = max_value, 
           label = paste("Pseudo R² = ", 
                         signif(adj.r.squared, 
                                digits = 3))),
           hjust = 0, 
           vjust = 1.5,
           size = 2)

@@

journal_title_size <- 2

@@

# Plot the PCA results
plot_pca_means <- 
ggplot() + 
  labs(x = paste0("PC1 (", variance_explained[1], "%)"),
       y = paste0("PC2 (", variance_explained[2], "%)")) +
  geom_point(data = pca_scores_means,
               aes(.fittedPC1,
                   .fittedPC2),
             size = 1) +
  geom_text_repel(data = pca_scores_means,
            aes(.fittedPC1,
                .fittedPC2,
                label = journal),
            lineheight = 0.8,
            segment.color = NA,
            size = 1,
            bg.color = "white",  # Color of the halo
            bg.r = 0.2) +
     geom_segment(data = pca_loadings_means, 
               aes(x = 0, 
                   y = 0, 
                   xend = PC1, 
                   yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "grey70") +
  geom_text_repel(data = pca_loadings_means,
            aes(PC1,
                PC2,
                label = column),
            size = 1,
            lineheight = 0.7,
            force = 0,
            segment.color = NA,
            color = "grey40",
           bg.color = "white",  # Color of the halo
          bg.r = 0.2) +
  theme_minimal(base_size = base_size) +
  coord_equal()

@@

plot_borda_scores <- 
  borda_scores %>% 
  mutate(Journal = str_wrap(Journal, 20)) %>% 
ggplot() +
  aes(reorder(Journal, Score),
      Score) +
  geom_col() +
  coord_flip() +
  ylab("Borda Count scores") + 
  xlab("") +
  theme_minimal(base_size = base_size) +
  theme(axis.text.y = element_text(size = journal_title_size))

@@

plot_journals_authors <- 
journal_metrics_for_plotting %>% 
  mutate(journal = str_wrap(journal, 20)) %>% 
  ggplot(aes(y = reorder(journal, 
                         log_authors,
                     FUN = mean),
             x = log_authors,
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal(base_size = base_size) +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  ylab("") + 
  xlab("Number of authors (ln)")

plot_journals_article_length <- 
journal_metrics_for_plotting %>% 
  mutate(journal = str_wrap(journal, 20)) %>% 
  ggplot(aes(y = reorder(journal,
                     -log_pages,
                     FUN = mean), 
             x = log_pages,
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal(base_size = base_size) +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  xlab("Number of pages (ln)") +
  ylab("")

plot_journals_title_length <- 
journal_metrics_for_plotting %>% 
  mutate(journal = str_wrap(journal, 20)) %>% 
  ggplot(aes(y = reorder(journal,
                         relative_title_length,
                         FUN = mean), 
             x = relative_title_length,
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal(base_size = base_size) +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  ylab("") +
  xlab("Relative title length (ln)")

plot_journals_ref_recency <- 
journal_metrics_for_plotting %>% 
  mutate(journal = str_wrap(journal, 20)) %>% 
  group_by(journal) %>% 
  ggplot(aes(y = reorder(journal,
                     prices_index,
                     FUN = mean), 
             x = prices_index,            
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal(base_size = base_size) +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  ylab("") +
  xlab("Recency of references")

plot_journals_ref_diversity <- 
journal_metrics_for_plotting %>%  
  mutate(journal = str_wrap(journal, 20)) %>% 
  group_by(journal) %>% 
  ggplot(aes(y = reorder(journal,
                     -shannon,
                     FUN = mean), 
             x = shannon,
             fill =   after_stat(x),
             height = after_stat(density))) +
  geom_density_ridges_gradient(stat = "density",
                               colour = "white") +
  scale_fill_viridis_c() +
  guides(fill = 'none') +
  theme_minimal(base_size = base_size) +
  theme(axis.text.y = element_text(size = journal_title_size)) +
  ylab("") +
  xlab("Diversity of references")
```

```{r}
#| label: fig-variation-by-journal
#| fig-cap: "Panels A-E: Variation in bibliometric indicators of hardness for 20 archaeological journals. The journals are ordered for each indicator so that within each plot, the harder journals are at the top of the plot and the softer journals are at the base. Panel F shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm. Panel G is a biplot of the first and second principal components of a PCA computed on the journal means of the five bibliometric variables."
#| fig-height: 6

@@

plot_grid(
  plot_grid(plot_journals_authors, 
           plot_journals_article_length,
           plot_journals_title_length,
           plot_journals_ref_recency, 
           plot_journals_ref_diversity,
           plot_borda_scores,
           nrow = 2,
           labels = LETTERS[1:6],
           label_size = 6),
  plot_pca_means,
  nrow = 2,
  labels = c("", LETTERS[7]),
  rel_heights = c(1, 0.5),
  label_size = 6)

@@

base_size <- 4

@@

# relationship between code issues and language?
aer_plot_issues_lang <- 
jas_rr_data %>% 
  separate_longer_delim(`Code issue`, ",") %>% 
  separate_longer_delim(Language, ",") %>% 
  mutate(Language = str_squish(Language)) %>% 
  mutate(`Code issue` = str_squish(`Code issue`)) %>% 
  group_by(`Code issue`, 
           Language) %>% 
  tally() %>% 
  drop_na() %>% 
  ggplot() +
  aes(`Code issue`, 
      Language,
      size = n) +
  geom_point() +
  xlab("Problem with the code") +
  ylab("Software") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  theme_minimal(base_size = base_size) +
  scale_size_area(breaks = c(1, 2, 3, 5)) +
  theme(legend.position = c(.9,.7),
        legend.title = element_text(hjust = 0.5),
        legend.spacing = unit(0.1, "cm"),  # Reduce space between legend items
        legend.margin = margin(1, 5, 1, 1),  # Reduce margin around the legend
        legend.key.size = unit(0.2, "cm"),  # Reduce size of legend keys
        legend.box.background=element_rect(fill="white", 
                                          color="black")) 
```

```{r}
#| label: fig-aer-summary
#| fig-cap: "Summary of reproducibility reviews for JAS. A: Primary software used for the computational analysis reported in a manuscript. B: Computational or statistical method used by the authors (GMM = geometric morphometrics, Frequentist = hypothesis tests such as chi-square and ANOVA, AI/ML = artifical intellegence and machine learning, including neural networks and deep learning, Network = statitical analysis of social networks, 3D = analysis of 3D data such as artefact models, Composition = compositional analysis of artefacts). C: Locations where authors deposited their code and data files. D: Issues that prevented the reproducible review from succeeding on the first attempt. E: Relationship between software used and issues that make research irreproducible."
#| fig-height: 3
